---
title: "Introduction to qml: Quantum Machine Learning"
author: "Xopre Rodríguez Gallego"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to qml: Quantum Machine Learning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(qml)
library(dplyr)
```

## Overview

The `qml` package provides a modern implementation of quantum matrix classification using the Density Matrix (DM) algorithm. This vignette introduces you to the main functions and demonstrates how to use them for classification tasks.

## Key Features

- **Quantum Matrix Classification**: Novel approach using quantum density matrices
- **Standard R Interface**: Compatible with `predict()`, `summary()`, and other R conventions
- **S3 Methods**: Professional implementation with proper documentation
- **Stratified Data Splitting**: Built-in support for balanced train/test splits

## Basic Usage

### Fitting a Model

The main function for fitting a quantum matrix classifier is `dm_fit()`. It uses a formula interface similar to other R modeling functions:

```{r basic-fit}
# Fit a model using the iris dataset
model <- dm_fit(Species ~ ., data = iris, n_breaks = 3)
print(model)
```

### Making Predictions

Use the standard `predict()` function to make predictions:

```{r predictions}
# Class predictions
predictions <- predict(model, newdata = iris[1:10, ])
print(predictions)

# Probability predictions
probabilities <- predict(model, newdata = iris[1:5, ], type = "prob")
print(probabilities)
```

### Model Summary

Get detailed information about your fitted model:

```{r summary}
summary(model)
```

## Data Splitting

The package includes the `testify()` function for creating stratified train/test splits:

```{r data-split}
# Create a stratified split
iris_split <- testify(iris, "Species", test_prob = 0.3)

# Check the split
table(iris_split$test, iris_split$Species)

# Separate into training and test sets
train_data <- iris_split[!iris_split$test, ]
test_data <- iris_split[iris_split$test, ]

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
```

## Advanced Usage

### Custom Parameters

You can customize the classifier behavior:

```{r custom-params}
# Use more discretization bins
model_custom <- dm_fit(
  Species ~ ., 
  data = train_data,
  n_breaks = 5,
  verbose = 1
)
```

### Model Evaluation

```{r evaluation}
# Fit on training data
model_eval <- dm_fit(Species ~ ., data = train_data, n_breaks = 3)

# Predict on test data
test_predictions <- predict(model_eval, newdata = test_data)

# Create confusion matrix
conf_matrix <- table(Predicted = test_predictions, Actual = test_data$Species)
print(conf_matrix)

# Calculate accuracy
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
cat("Accuracy:", round(accuracy * 100, 2), "%\n")
```

## Algorithm Overview

The quantum matrix classifier works through several key steps:

1. **Data Preprocessing**: Discretizes continuous variables and creates quantum state representations
2. **Quantum Matrix Construction**: Builds quantum density matrices from the preprocessed data
3. **Coordinate Transformation**: Projects data onto eigenvector basis using SVD
4. **Polar Coordinates**: Converts to polar coordinate system for classification
5. **Kernel Density Estimation**: Uses quantum kernels for probability estimation
6. **Classification**: Applies maximum likelihood for final predictions

## Mathematical Functions

The package also exports key mathematical functions that you can use directly:

```{r math-functions}
# Quantum kernel function
K(0.1, 0.2, r = 2)

# The kernel is 0 outside the unit circle
K(1.1, 0.5, r = 2)
```

## Conclusion

The `qml` package provides a modern, well-structured implementation of quantum matrix classification. Its standard R interface makes it easy to integrate into existing machine learning workflows, while the modular architecture ensures maintainability and extensibility.
