---
title: "Density Matrix Model: Hard Case"
author: "Xopre Rodríguez Gallego"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
codification: "UTF-8"
output:
  ioslides_presentation:
    widescreen: true
    transition: slower
    css: ../config/style/matteocourthoud_custom_ioslides.css
    logo: ../config/style/logo.svg
params:
  test_prop: 0.25
  random_seed: 42
---

```{r setup, include=FALSE}
source(here::here("config/setup.R"))
set.seed(params$random_seed)

library(qml)
library(dplyr)
library(ggplot2)
library(rsample)
library(yardstick)
```

# Harder Binary Classification

## Task overview

In this streamlined example we tackle a harder binary classification task using the classic `iris` dataset. We remove the easily separable `setosa` class and keep only `versicolor` and `virginica`, which are known to be more difficult to distinguish.

- Dataset: `datasets::iris`
- Objective: classify `versicolor` vs `virginica`
- Features: all remaining numeric variables

This mirrors the structure used in the DM slides: short context + focused code chunks.

## Data preparation

```{r DMH-1}
Data <- datasets::iris %>%
  dplyr::filter(Species %in% c("versicolor", "virginica")) %>%
  dplyr::mutate(
    y = ifelse(Species == "versicolor", "0", "1"),
    .keep = "unused"
  ) %>%
  dplyr::mutate(y = factor(y)) %>%
  dplyr::relocate(y)

# quick glance
Data %>% dplyr::glimpse()
```

## Visual overlap (why this is hard)

```{r DMH-2, echo=FALSE}
# A quick scatter to show overlap between classes
Data %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = y)) +
  geom_point(alpha = 0.8) +
  scale_color_brewer(palette = "Set1", name = "Class") +
  labs(title = "Iris: versicolor (0) vs virginica (1)",
       subtitle = "Harder boundary than setosa vs others",
       x = "Petal Length", y = "Petal Width") +
  theme_minimal()
```

Reading the plot:

- Points from the two classes overlap noticeably in this 2D view, which anticipates a tougher decision boundary.
- Expect lower separation than in the easy penguins example; confidence may be more spread (see probability densities later).

## Train/Test split

```{r DMH-3}
Data_split <- initial_split(Data, prop = 1 - params$test_prop, strata = "y")
Data_train <- training(Data_split)
Data_test  <- testing(Data_split)

# class balance check
rbind(
  train = table(Data_train$y),
  test  = table(Data_test$y)
)
```

# Fit and Predict

## Fit DM model (modern interface)

```{r DMH-4}
# Fit the DM classifier (polar coords by default)
dm_model <- dm_fit(y ~ ., data = Data_train, n_breaks = 3)
print(dm_model)
```

Notes:

- Follows the modern `dm_fit()`/`predict()` convention, mirroring the DM deck.
- Stores preprocessing artifacts to keep prediction transformations consistent.

## Predictions

```{r DMH-5}
class_pred <- predict(dm_model, newdata = Data_test)
prob_pred  <- predict(dm_model, newdata = Data_test, type = "prob")

# Combine for analysis
result <- Data_test %>%
  dplyr::mutate(
    .pred  = class_pred,
    .truth = y
  ) %>%
  dplyr::bind_cols(prob_pred)

head(result)
```

# Evaluation

## Metrics

```{r DMH-6}
my_metrics <- metric_set(bal_accuracy, accuracy, sensitivity, specificity, f_meas, kap, mcc)
my_metrics(result, truth = .truth, estimate = .pred)
```

How to read:

- Balanced accuracy is robust to class imbalance; higher is better.
- Sensitivity/specificity show trade-offs by class; kappa and MCC summarize agreement vs chance.

## ROC AUC

```{r DMH-7}
roc_auc(result, truth = .truth, .pred_1, event_level = "second")
```

Interpretation: AUC close to 1 indicates good ranking of positives vs negatives; ~0.5 means no better than random.

## Confusion matrix

```{r DMH-8, message=FALSE}
cm <- conf_mat(data = result, truth = .truth, estimate = .pred)
autoplot(cm, type = "heatmap") + theme_minimal()
```

Diagonal cells are correct predictions; off-diagonal are errors. Compare magnitudes to gauge which class is harder.

## Probability distributions (confidence)

```{r DMH-9, echo=FALSE}
result %>%
  dplyr::select(starts_with(".pred_"), .truth) %>%
  tidyr::pivot_longer(cols = starts_with(".pred_"),
                      names_to = "predicted_class",
                      values_to = "probability") %>%
  dplyr::mutate(predicted_class = stringr::str_remove(predicted_class, "^\\.pred_")) %>%
  ggplot(aes(x = probability, fill = predicted_class)) +
  geom_density(alpha = 0.7, adjust = 1) +
  facet_wrap(~.truth, labeller = label_both) +
  labs(title = "Prediction Probability Distribution by True Class",
       x = "Predicted Probability", y = "Density",
       fill = "Predicted Class") +
  theme_minimal()
```

Guidance (as in DM):

- Within the facet of the true class, the corresponding class curve should peak near 1.
- Overlap between curves reflects uncertainty; strong overlap suggests limited separability in this task.

# Optional: Cartesian coordinates

If desired, you can try the model without polar coordinates (first two eigen-basis components). Toggle `use_polar = FALSE`:

```{r DMH-10, eval=FALSE}
dm_model_xy <- dm_fit(y ~ ., data = Data_train, n_breaks = 3, use_polar = FALSE)
class_pred_xy <- predict(dm_model_xy, newdata = Data_test)
prob_pred_xy  <- predict(dm_model_xy, newdata = Data_test, type = "prob")
```

This variant removes polar conversion and uses the first two eigen-basis coordinates. It can help isolate where gains come from.

# Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
