---
title: "DM vs Plain Kernel Density (f_hat)"
author: "Xopre Rodríguez Gallego"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
codification: "UTF-8"
output:
  ioslides_presentation:
    widescreen: true
    transition: slower
    css: ../config/style/matteocourthoud_custom_ioslides.css
    logo: ../config/style/logo.svg
params:
  test_prop: 0.25
  random_seed: 42
  baseline_features: ["Petal.Length", "Petal.Width"]
---

```{r setup, include=FALSE}
source(here::here("config/setup.R"))
set.seed(params$random_seed)

library(qml)
library(dplyr)
library(ggplot2)
library(rsample)
library(yardstick)
library(tidyr)
library(purrr)

# Helper: get f_hat_h_cartesian (from qml if available, else fallback)
if (!exists("f_hat_h_cartesian")) {
  f_hat_h_cartesian <- function(D_tilde_xy, x, clase, h_window = 0.1) {
    n_total <- D_tilde_xy %>% dplyr::bind_rows() %>% nrow()
    D_tilde_xy[[clase]] %>%
      dplyr::mutate(
        v_1 = (x[1] - Coord_1) / h_window,
        v_2 = (x[2] - Coord_2) / h_window
      ) %>%
      dplyr::mutate(kernel_value = purrr::pmap_dbl(dplyr::select(., v_1, v_2), qml::K)) %>%
      dplyr::pull(kernel_value) %>%
      sum() %>%
      magrittr::divide_by(n_total * h_window^2)
  }
}
```

# Goal and Design

## What we test

- DM pipeline: change coordinates via the quantum matrix (as in `dm_fit()`/`predict()`), then density-based classification.
- Plain kernel density baseline: apply the same kernel density classification directly on the raw feature space (2D), without quantum coordinate changes.

Both use the same train/test split and the same bandwidth to be comparable.

Style note (as in DM): short context + focused chunks, with a quick “how to read” where helpful.

# Data and split

## Dataset

We reuse a harder binary problem as in DM_hard: `versicolor` vs `virginica` from `datasets::iris`.

```{r DMC-1}
Data <- datasets::iris %>%
  dplyr::filter(Species %in% c("versicolor", "virginica")) %>%
  dplyr::mutate(
    y = ifelse(Species == "versicolor", "0", "1"),
    .keep = "unused"
  ) %>%
  dplyr::mutate(y = factor(y)) %>%
  dplyr::relocate(y)

Data %>% dplyr::glimpse()
```

Why this dataset: Versicolor vs Virginica presents a more overlapping boundary than Setosa vs others, making it a fair stress test for the coordinate transform.

Visual overlap on the chosen baseline features (for intuition):

```{r DMC-1b, echo=FALSE}
raw_feats <- params$baseline_features
Data %>%
  ggplot(aes_string(x = raw_feats[1], y = raw_feats[2], color = "y")) +
  geom_point(alpha = 0.8) +
  scale_color_brewer(palette = "Set1", name = "Class") +
  labs(title = paste("Iris:", paste(raw_feats, collapse = " vs ")), subtitle = "Baseline 2D space used by the kernel",
       x = raw_feats[1], y = raw_feats[2]) +
  theme_minimal()
```

## Train/Test split

```{r DMC-2}
Data_split <- initial_split(Data, prop = 1 - params$test_prop, strata = "y")
Data_train <- training(Data_split)
Data_test  <- testing(Data_split)

rbind(
  train = table(Data_train$y),
  test  = table(Data_test$y)
)
```

# Model A: DM pipeline

## Fit and predict with DM

```{r DMC-3}
# Fit DM classifier (polar by default)
dm_model <- dm_fit(y ~ ., data = Data_train, n_breaks = 3)

# Use model bandwidth for the baseline, too
common_bandwidth <- dm_model$bandwidth

# Predictions
pred_class_dm <- predict(dm_model, newdata = Data_test)
pred_prob_dm  <- predict(dm_model, newdata = Data_test, type = "prob")

result_dm <- Data_test %>%
  dplyr::mutate(
    .pred  = pred_class_dm,
    .truth = y
  ) %>%
  dplyr::bind_cols(pred_prob_dm)
head(result_dm)
```

As in DM slides, we keep the preview minimal—just enough to verify the structure.

# Model B: Plain kernel density on raw features

## Build baseline training data in 2D

We choose two raw features (default: Petal.Length, Petal.Width), standardize using the training set, and classify via the same kernel.

```{r DMC-4}
raw_feats <- params$baseline_features
stopifnot(all(raw_feats %in% names(Data_train)))

# Standardization fit on train
mu <- sapply(Data_train[raw_feats], mean)
sd <- sapply(Data_train[raw_feats], sd)

std <- function(df) {
  as_tibble(scale(df[raw_feats], center = mu, scale = sd)) %>%
    setNames(c("Coord_1", "Coord_2"))
}

train_std <- std(Data_train) %>% dplyr::mutate(y = Data_train$y)

# Split into class lists with Coord_1, Coord_2 columns
baseline_train_list <- list(
  train_std %>% dplyr::filter(y == "0") %>% dplyr::select(Coord_1, Coord_2),
  train_std %>% dplyr::filter(y == "1") %>% dplyr::select(Coord_1, Coord_2)
)
```

Notes:

- Standardization uses train mean/sd and is applied to test—mirrors typical modeling practice.
- Naming columns `Coord_1` and `Coord_2` lets us reuse the same kernel machinery.

## Predict on test with baseline

```{r DMC-5}
# Standardize test with train parameters
test_std <- std(Data_test)

# For each test row, compute class densities and form predictions
predict_baseline <- function(x1, x2, bandwidth) {
  probs <- seq_along(baseline_train_list) %>%
    map_dbl(function(cls) f_hat_h_cartesian(baseline_train_list, c(x1, x2), clase = cls, h_window = bandwidth))
  if (all(probs == 0)) {
    probs <- rep(1/length(baseline_train_list), length(baseline_train_list))
  } else {
    probs <- probs / sum(probs)
  }
  list(class = which.max(probs), probs = probs)
}

pred_list <- pmap(test_std, ~ predict_baseline(..1, ..2, common_bandwidth))

pred_class_base <- map_int(pred_list, ~ .x$class) %>% factor(levels = 1:2, labels = c("0","1"))
pred_prob_base  <- map(pred_list, ~ .x$probs) %>% do.call(rbind, .) %>% as_tibble()
colnames(pred_prob_base) <- c(".pred_0", ".pred_1")

result_base <- Data_test %>%
  dplyr::mutate(
    .pred  = pred_class_base,
    .truth = y
  ) %>%
  dplyr::bind_cols(pred_prob_base)
head(result_base)
```

This forms a like-for-like density estimator on the raw 2D space with the same bandwidth.

# Comparison

## Metrics side-by-side

```{r DMC-6}
met <- metric_set(bal_accuracy, accuracy, sensitivity, specificity, f_meas, kap, mcc)

m_dm <- met(result_dm, truth = .truth, estimate = .pred) %>% dplyr::mutate(model = "DM")
m_ba <- met(result_base, truth = .truth, estimate = .pred) %>% dplyr::mutate(model = "Baseline")

bind_rows(m_dm, m_ba) %>%
  tidyr::pivot_wider(names_from = model, values_from = .estimate) %>%
  dplyr::arrange(.metric) %>%
  knitr::kable(caption = "Metrics: DM vs Plain Kernel Density Baseline")
```

How to read (following DM):

- Higher balanced accuracy indicates better average performance across classes.
- Kappa/MCC capture agreement beyond chance; useful for nuanced comparisons.

## ROC AUC

```{r DMC-7}
auc_dm <- roc_auc(result_dm, truth = .truth, .pred_1, event_level = "second") %>% dplyr::mutate(model = "DM")
auc_ba <- roc_auc(result_base, truth = .truth, .pred_1, event_level = "second") %>% dplyr::mutate(model = "Baseline")

bind_rows(auc_dm, auc_ba) %>% knitr::kable(caption = "ROC AUC: DM vs Baseline")
```

If DM’s coordinate change adds value, we expect a consistently higher AUC than the baseline.

## Confusion matrices

```{r DMC-8, message=FALSE}
cm_dm  <- conf_mat(result_dm, truth = .truth, estimate = .pred)
cm_ba  <- conf_mat(result_base, truth = .truth, estimate = .pred)

list(DM = cm_dm, Baseline = cm_ba)
```

Diagonal dominance indicates better classification; compare error types across methods.

## Probability distributions

```{r DMC-9, echo=FALSE}
plot_probs <- function(df, title) {
  df %>%
    dplyr::select(starts_with(".pred_"), .truth) %>%
    tidyr::pivot_longer(cols = starts_with(".pred_"), names_to = "predicted_class", values_to = "probability") %>%
    dplyr::mutate(predicted_class = stringr::str_remove(predicted_class, "^\\.pred_")) %>%
    ggplot(aes(x = probability, fill = predicted_class)) +
    geom_density(alpha = 0.7) +
    facet_wrap(~.truth, labeller = label_both) +
    labs(title = title, x = "Predicted Probability", y = "Density", fill = "Predicted Class") +
    theme_minimal()
}

plot_probs(result_dm,  "DM: Probability Distributions")
plot_probs(result_base,"Baseline: Probability Distributions")
```

Reading guidance (as in DM):

- Within each facet (true class), the corresponding class curve should peak near probability 1.
- Less overlap and sharper peaks generally indicate a more confident, better-calibrated model.

# Notes

- Same split, same bandwidth: fair comparison.
- Differences reflect the added structure from the quantum coordinate transform vs working directly in raw feature space.

Takeaways:

- If DM outperforms the baseline across metrics and AUC, the quantum-induced coordinate change is helpful on this task.
- If results are similar, the raw 2D space may be adequate, or bandwidth/feature choices may need tuning.

# Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
