---
title: "DM vs ML: Hard Case"
author: "Xopre Rodríguez Gallego"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
codification: "UTF-8"
output:
  ioslides_presentation:
    widescreen: true
    transition: slower
    css: ../config/style/matteocourthoud_custom_ioslides.css
    logo: ../config/style/logo.svg
params:
  test_prop: 0.25
  random_seed: 42
---

```{r setup, include=FALSE}
source(here::here("config/setup.R"))
set.seed(params$random_seed)

library(qml)
library(dplyr)
library(ggplot2)
library(rsample)
library(yardstick)
library(recipes)
library(parsnip)
library(workflows)
```

# Harder Binary Classification

## Task overview

We focus on the hard iris task: distinguishing `versicolor` from `virginica` (the separability is much lower than with `setosa`). This deck mirrors the concise style from the DM slides: short context, clean code, and targeted visuals.

- Dataset: `datasets::iris`
- Objective: classify `versicolor` vs `virginica`
- Features: all remaining numeric variables

## Data preparation

```{r DMnML-1}
Data <- datasets::iris %>%
  dplyr::filter(Species %in% c("versicolor", "virginica")) %>%
  dplyr::mutate(
    y = ifelse(Species == "versicolor", "0", "1"),
    .keep = "unused"
  ) %>%
  dplyr::mutate(y = factor(y)) %>%
  dplyr::relocate(y)

# quick glance
Data %>% dplyr::glimpse()
```

## Visual overlap (why this is hard)

```{r DMnML-2, echo=FALSE}
Data %>%
  ggplot(aes(x = Petal.Length, y = Petal.Width, color = y)) +
  geom_point(alpha = 0.8) +
  scale_color_brewer(palette = "Set1", name = "Class") +
  labs(title = "Iris: versicolor (0) vs virginica (1)",
       subtitle = "Noticeable overlap → tougher boundary",
       x = "Petal Length", y = "Petal Width") +
  theme_minimal()
```

Reading the plot:

- Classes overlap considerably in this 2D view; expect more ambiguous probabilities than in the easy case.
- We'll compare DM to simple ML baselines using the same split and metrics.

## Train/Test split

```{r DMnML-3}
Data_split <- initial_split(Data, prop = 1 - params$test_prop, strata = "y")
Data_train <- training(Data_split)
Data_test  <- testing(Data_split)

# class balance check
rbind(
  train = table(Data_train$y),
  test  = table(Data_test$y)
)
```

# DM model

## Fit and predict (polar by default)

```{r DMnML-4}
dm_model <- dm_fit(y ~ ., data = Data_train, n_breaks = 3)

pred_class_dm <- predict(dm_model, newdata = Data_test)
pred_prob_dm  <- predict(dm_model, newdata = Data_test, type = "prob")

result_dm <- Data_test %>%
  dplyr::mutate(
    .pred  = pred_class_dm,
    .truth = y
  ) %>%
  dplyr::bind_cols(pred_prob_dm)

head(result_dm)
```

## Metrics and diagnostics

```{r DMnML-5}
my_metrics <- metric_set(bal_accuracy, accuracy, sensitivity, specificity, f_meas, kap, mcc)
my_metrics(result_dm, truth = .truth, estimate = .pred)
```

```{r DMnML-6}
roc_auc(result_dm, truth = .truth, .pred_1, event_level = "second")
```

```{r DMnML-7, message=FALSE}
cm_dm <- conf_mat(data = result_dm, truth = .truth, estimate = .pred)
autoplot(cm_dm, type = "heatmap") + theme_minimal()
```

How to read:

- Balanced accuracy summarizes class-wise performance; kappa/MCC account for chance agreement.
- AUC shows ranking quality; confusion heatmap highlights error patterns.

# ML baselines

## Simple recipe and models

We keep baselines lightweight and comparable. Numeric standardization helps distance-based models.

```{r DMnML-8}
ml_rec <- recipe(y ~ ., data = Data_train) %>%
  step_zv(all_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

# Model specs
log_spec <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

knn_spec <- nearest_neighbor(neighbors = 11) %>%
  set_engine("kknn") %>%
  set_mode("classification")

rf_spec <- rand_forest(trees = 500, mtry = 2, min_n = 5) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# Workflows
wf_log <- workflow() %>% add_model(log_spec) %>% add_recipe(ml_rec)
wf_knn <- workflow() %>% add_model(knn_spec) %>% add_recipe(ml_rec)
wf_rf  <- workflow() %>% add_model(rf_spec)  %>% add_recipe(ml_rec)

fit_log <- fit(wf_log, Data_train)
fit_knn <- fit(wf_knn, Data_train)
fit_rf  <- fit(wf_rf,  Data_train)
```

## Predictions on test

```{r DMnML-9}
predict_with_probs <- function(fitted, new_data) {
  tibble(
    .pred   = predict(fitted, new_data) %>% dplyr::pull(1),
    .truth  = new_data$y
  ) %>%
    dplyr::bind_cols(predict(fitted, new_data, type = "prob"))
}

res_log <- predict_with_probs(fit_log, Data_test)
res_knn <- predict_with_probs(fit_knn, Data_test)
res_rf  <- predict_with_probs(fit_rf,  Data_test)
```

## Metrics table (test set)

```{r DMnML-10}
summarize_metrics <- function(df, model_name) {
  tibble(model = model_name) %>%
    dplyr::bind_cols(
      my_metrics(df, truth = .truth, estimate = .pred) %>%
        dplyr::select(.metric, .estimate) %>%
        tidyr::pivot_wider(names_from = .metric, values_from = .estimate)
    ) %>%
    dplyr::mutate(
      auc = roc_auc(df, truth = .truth, .pred_1, event_level = "second")$.estimate
    )
}

metrics_dm  <- summarize_metrics(result_dm, "DM")
metrics_log <- summarize_metrics(res_log,   "Logistic")
metrics_knn <- summarize_metrics(res_knn,   "KNN")
metrics_rf  <- summarize_metrics(res_rf,    "RF")

metrics_all <- dplyr::bind_rows(metrics_dm, metrics_log, metrics_knn, metrics_rf) %>%
  dplyr::mutate(dplyr::across(-model, round, 3)) %>%
  dplyr::arrange(dplyr::desc(bal_accuracy))

knitr::kable(metrics_all)
```

Notes:

- Same split and metrics across DM and ML ensure a fair comparison on this harder task.
- If desired, you can extend with more models or cross-validation, but this keeps the focus on clarity.

# Probability distributions (confidence)

```{r DMnML-11, echo=FALSE}
result_dm %>%
  dplyr::select(starts_with(".pred_"), .truth) %>%
  tidyr::pivot_longer(cols = starts_with(".pred_"),
                      names_to = "predicted_class",
                      values_to = "probability") %>%
  dplyr::mutate(predicted_class = stringr::str_remove(predicted_class, "^\\.pred_")) %>%
  ggplot(aes(x = probability, fill = predicted_class)) +
  geom_density(alpha = 0.7, adjust = 1) +
  facet_wrap(~.truth, labeller = label_both) +
  labs(title = "DM: Prediction Probability Distribution by True Class",
       x = "Predicted Probability", y = "Density",
       fill = "Predicted Class") +
  theme_minimal()
```

Guidance:

- Within each true-class facet, the matching predicted-class density ideally peaks near 1.
- Overlap indicates uncertainty; heavier overlap is expected here versus easier problems.

# Optional: Cartesian DM variant

```{r DMnML-12, eval=FALSE}
dm_xy <- dm_fit(y ~ ., data = Data_train, n_breaks = 3, use_polar = FALSE)
res_xy <- Data_test %>%
  dplyr::mutate(
    .pred  = predict(dm_xy, newdata = Data_test),
    .truth = y
  ) %>%
  dplyr::bind_cols(predict(dm_xy, newdata = Data_test, type = "prob"))

my_metrics(res_xy, truth = .truth, estimate = .pred)
roc_auc(res_xy, truth = .truth, .pred_1, event_level = "second")
```

# Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
