---
title: "Density Matrix vs Principal Component Analysis"
subtitle: "A Comparative Study of Dimensionality Reduction Methods"
author: "Xopre Rodríguez Gallego"
codification: "UTF-8"
output: 
  ioslides_presentation:
    widescreen: true # Wider form factor
    # smaller: true # Text
    transition: slower # default
    css: ../config/style/matteocourthoud_custom_ioslides.css
    logo: ../config/style/logo.svg
params:
  test_prop: 0.3
  n_components: 2
  n_breaks: 3
  random_seed: 42
---

```{r setup, include=FALSE}
source(here::here("config/setup.R"))

# {
#   devtools::document()
#   devtools::install(".", upgrade = "never", quiet = TRUE)
#   cat("✓ Package installed successfully\n")
# }
library(qml)  # Modern qml package with dm_fit/predict interface
# Print package version
cat("Using qml version:", as.character(packageVersion("qml")), "\n")
```


```{r DMnPCA-1, include=FALSE}
library(tidymodels)
library(yardstick)
library(patchwork)

# Set random seed for reproducibility
set.seed(params$random_seed)
```

# Introduction

## Dimensionality Reduction: Two Paradigms

**Objective**: Compare two fundamentally different approaches to dimensionality reduction:

# Goal

Focus on how coordinates and variance change under PCA vs DM, and add projection-quality metrics. We’ll: (1) visualize the new coordinates, (2) quantify variance “before/after”, and (3) derive simple metrics to understand what each projection emphasizes.

```{r}
# Load and examine the dataset
data <- modeldata::penguins %>%
  filter(complete.cases(.)) %>%  # Remove missing values
# We'll use the Palmer penguins dataset (Adelie vs Gentoo) to compare the projections.
  mutate(
    species = droplevels(species),
    species = factor(species, levels = c("Adelie", "Gentoo"))
  ) %>% 
  na.omit()

glimpse(data)
```

**Dataset characteristics:**
- **Binary classification**: Adelie vs Gentoo penguins
- **`r nrow(data)` observations** with `r ncol(data)-1` predictor variables
- **Mixed data types**: Continuous and categorical variables

# Data Exploration

## Dataset Overview

```{r DMnPCA-2, echo=FALSE, fig.height=5}
# Visualize the raw data
# Data split and “before” variance
data_split <- initial_split(data, prop = 1 - params$test_prop, strata = species)
train_data <- training(data_split)
test_data <- testing(data_split)

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
cat("Class balance in training:")
table(train_data$species)
```

# Method 1: Principal Component Analysis (PCA)

## PCA: Classical Approach

**Principal Component Analysis (PCA)**:

```{r DMnPCA-3 }
# Compute variance in original numeric space (standardized), training only
num_train <- train_data %>% dplyr::select(where(is.numeric))
num_train_scaled <- num_train %>% mutate(across(everything(), scale))
orig_total_var <- num_train_scaled %>% summarise(across(everything(), ~ var(.x, na.rm = TRUE))) %>%
  rowSums(na.rm = TRUE)
cat("Original (standardized) total variance (train, numeric only):", round(orig_total_var, 3), "\n")
```
- **Linear transformation** that maximizes variance
- Finds orthogonal directions (principal components)
- **Assumption**: Linear relationships capture the important structure
- **Output**: Coordinates in PC space

**Mathematical Foundation:**
$$\mathbf{Y} = \mathbf{X} \mathbf{W}$$
where $\mathbf{W}$ contains the eigenvectors of the covariance matrix.

## PCA Implementation

```{r DMnPCA-4 }
# Prepare data for PCA (numeric variables only)
pca_recipe <- recipe(species ~ ., data = train_data) %>%
  step_select(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_pca(all_numeric_predictors(), num_comp = params$n_components)

# Fit PCA transformation
pca_prep <- prep(pca_recipe, training = train_data)

# Transform both training and test sets
train_pca <- bake(pca_prep, new_data = train_data)
test_pca <- bake(pca_prep, new_data = test_data)

# Extract PCA transformation details
pca_obj <- pca_prep$steps[[3]]  # PCA step
pca_loadings <- pca_obj$res$rotation[, 1:params$n_components]
pca_variance <- summary(pca_obj$res)$importance[2, 1:params$n_components]

cat("PCA Variance explained:")
print(pca_variance)
```

## PCA Visualization: Transformed Space

```{r DMnPCA-5, echo=FALSE, fig.height=5}
# Visualize PCA transformation
p1 <- train_pca %>%
  bind_cols(species = train_data$species) %>% 
  ggplot(aes(PC1, PC2, color = species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  labs(
    title = "PCA: Training Data in PC Space",
    x = paste0("PC1 (", round(pca_variance[1]*100, 1), "% variance)"),
    y = paste0("PC2 (", round(pca_variance[2]*100, 1), "% variance)")
  ) +
  theme_minimal()

p2 <- test_pca %>%
  bind_cols(species = test_data$species) %>% 
  ggplot(aes(PC1, PC2, color = species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  labs(
    title = "PCA: Test Data in PC Space",
    x = paste0("PC1 (", round(pca_variance[1]*100, 1), "% variance)"),
    y = paste0("PC2 (", round(pca_variance[2]*100, 1), "% variance)")
  ) +
  theme_minimal()

p1 + p2 + plot_layout(guides = "collect")
```

## PCA: Feature Loadings

```{r DMnPCA-6, echo=FALSE}
# Visualize PCA loadings
pca_loadings %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  pivot_longer(-feature, names_to = "component", values_to = "loading") %>%
  ggplot(aes(feature, loading, fill = component)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "PCA Feature Loadings",
    x = "Original Features",
    y = "Loading Coefficient"
  ) +
  theme_minimal()
```

# Method 2: Density Matrix (DM)

## DM: Quantum-Inspired Approach

**Density Matrix Method**:
- **Quantum-inspired transformation** via density matrices
- Creates non-linear coordinate system
- **Converts to polar coordinates**: $(r, \phi)$
- **No linearity assumption** - captures complex relationships

**Mathematical Foundation:**
$$\rho_d = \frac{XX^T}{\text{tr}(XX^T)}$$
Followed by SVD and coordinate transformation to polar space.

## DM Implementation

```{r DMnPCA-7 }
# Fit DM model using the modern qml interface
dm_model <- dm_fit(species ~ ., data = train_data, n_breaks = params$n_breaks)

# Get model summary
print(dm_model)
```

## DM Implementation

```{r DMnPCA-8 }
# Extract training data characteristics
cat("\nTraining data characteristics:")
for(i in seq_along(dm_model$trained_data)) {
  cat("\nClass", i, "(", dm_model$classes[i], "):", 
      nrow(dm_model$trained_data[[i]]), "observations")
}
```

## DM Visualization: Polar Coordinate Space

```{r DMnPCA-9, echo=FALSE, fig.height=5}
# Create visualization data by extracting polar coordinates from trained model
# Use the NEW feature: return_coords = TRUE to get test data coordinates

# Get training data polar coordinates
train_dm_viz <- bind_rows(
  dm_model$trained_data[[1]] %>% mutate(species = dm_model$classes[1]),
  dm_model$trained_data[[2]] %>% mutate(species = dm_model$classes[2])
)

# NEW FEATURE: Get test data with polar coordinates
test_pred_with_coords <- predict(dm_model, newdata = test_data, return_coords = TRUE)

# Extract test coordinates and add actual class labels
test_dm_viz <- test_pred_with_coords$coords %>%
  bind_cols(species = test_data$species)

# Training data visualization
p1 <- train_dm_viz %>%
  ggplot(aes(r, phi_1, color = species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  coord_polar() +
  labs(
    title = "DM: Training Data in Polar Space",
    x = "r (radial distance)",
    y = "φ₁ (angle)"
  ) +
  theme_minimal()

# Test data visualization - NOW POSSIBLE with return_coords = TRUE!
p2 <- test_dm_viz %>%
  ggplot(aes(r, phi_1, color = species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  coord_polar() +
  labs(
    title = "DM: Test Data in Polar Space",
    x = "r (radial distance)", 
    y = "φ₁ (angle)"
  ) +
  theme_minimal()

p1 + p2 + plot_layout(guides = "collect")
```

## DM: Coordinate Space Properties

```{r DMnPCA-10, echo=FALSE}
# Analyze the polar coordinate distributions for both training and test
train_coord_stats <- train_dm_viz %>%
  group_by(species) %>%
  summarise(
    dataset = "Training",
    n = n(),
    r_mean = mean(r, na.rm = TRUE),
    r_sd = sd(r, na.rm = TRUE),
    phi_mean = mean(phi_1, na.rm = TRUE),
    phi_sd = sd(phi_1, na.rm = TRUE),
    .groups = "drop"
  )

test_coord_stats <- test_dm_viz %>%
  group_by(species) %>%
  summarise(
    dataset = "Test",
    n = n(),
    r_mean = mean(r, na.rm = TRUE),
    r_sd = sd(r, na.rm = TRUE),
    phi_mean = mean(phi_1, na.rm = TRUE),
    phi_sd = sd(phi_1, na.rm = TRUE),
    .groups = "drop"
  )

coord_stats <- bind_rows(train_coord_stats, test_coord_stats) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

coord_stats %>%
  knitr::kable(caption = "Polar Coordinate Statistics by Species and Dataset")
```

## NEW FEATURE: Polar Coordinates Visualization

**The `predict()` function now supports `return_coords = TRUE`:**

```{r DMnPCA-11, echo=TRUE}
# Standard prediction
pred_standard <- predict(dm_model, test_data, type = "class")

# NEW: Predictions with polar coordinates
pred_with_coords <- predict(dm_model, test_data, return_coords = TRUE)

cat("Standard prediction structure:\n")
str(pred_standard)

cat("\nWith coordinates structure:\n") 
str(pred_with_coords)

cat("\nFirst few polar coordinates:\n")
head(pred_with_coords$coords)
```

# Classification Performance

## Metrics Setup

```{r DMnPCA-12 }
# Define comprehensive metrics for comparison
classification_metrics <- metric_set(
  accuracy, bal_accuracy, sens, spec, 
  precision, recall, f_meas, kap, roc_auc
)
```

## PCA + Logistic Regression Performance

```{r DMnPCA-13 }
# Fit logistic regression on PCA-transformed data
pca_model <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(species ~ ., data = train_pca %>% bind_cols(species = train_data$species))

# Make predictions
pca_pred <- predict(pca_model, test_pca, type = "class")
pca_prob <- predict(pca_model, test_pca, type = "prob")

# Calculate metrics
# Session Info

