---
title: "Density Matrix vs Principal Component Analysis"
subtitle: "A Comparative Study of Dimensionality Reduction Methods"
author: "Xopre Rodríguez Gallego"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
codification: "UTF-8"
output: 
  ioslides_presentation:
    widescreen: true # Wider form factor
    # smaller: true # Text
    transition: slower # default
    css: ../config/style/matteocourthoud_custom_ioslides.css
    logo: ../config/style/logo.svg
params:
  test_prop: 0.3
  n_components: 2
  n_breaks: 3
  random_seed: 42
---

```{r setup, include=FALSE}
source(here::here("config/setup.R"))
library(tidyverse)

# {
#   devtools::document()
#   devtools::install(".", upgrade = "never", quiet = TRUE)
#   cat("✓ Package installed successfully\n")
# }
library(qml)  # Modern qml package with dm_fit/predict interface
# Print package version
cat("Using qml version:", as.character(packageVersion("qml")), "\n")
```


```{r setup-2, include=FALSE}
library(tidymodels)
library(yardstick)
library(patchwork)

# Set random seed for reproducibility
set.seed(params$random_seed)
```

# Introduction

## Dimensionality Reduction: Two Paradigms

**Objective**: Compare two fundamentally different approaches to dimensionality reduction:

1. **Principal Component Analysis (PCA)** - Classical linear method
   - Finds directions of maximum variance
   - Linear transformations
   - Orthogonal components

2. **Density Matrix (DM)** - Quantum-inspired method  
   - Creates quantum density matrices
   - Non-linear coordinate transformations
   - Polar coordinate representation

## Why Compare DM vs PCA?

**Key Questions:**
- How do quantum-inspired methods compare to classical approaches?
- When does non-linear transformation provide advantages?
- What are the computational and interpretability trade-offs?

**Expected Insights:**
- Performance comparison on classification tasks
- Visualization quality in reduced dimensions
- Robustness to data characteristics

## Case Study - Palmer Penguins

We'll use the Palmer penguins dataset to compare both methods on a real classification problem.

```{r load-data, echo=FALSE}
# Load and examine the dataset
data <- modeldata::penguins %>%
  filter(complete.cases(.)) %>%  # Remove missing values
  filter(species != "Chinstrap") %>%  # Binary classification
  mutate(
    species = droplevels(species),
    species = factor(species, levels = c("Adelie", "Gentoo"))
  )

glimpse(data)
```

**Dataset characteristics:**
- **Binary classification**: Adelie vs Gentoo penguins
- **`r nrow(data)` observations** with `r ncol(data)-1` predictor variables
- **Mixed data types**: Continuous and categorical variables

# Data Exploration

## Dataset Overview

```{r data-overview, echo=FALSE, fig.height=5}
# Visualize the raw data
p1 <- data %>%
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "Original Feature Space", 
       x = "Bill Length (mm)", y = "Bill Depth (mm)") +
  theme_minimal()

p2 <- data %>%
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(title = "Original Feature Space", 
       x = "Flipper Length (mm)", y = "Body Mass (g)") +
  theme_minimal()

p1 + p2 + plot_layout(guides = "collect")
```

## Class Distribution and Correlation

```{r class-distribution, echo=FALSE, fig.height=4}
# Class distribution
p1 <- data %>%
  ggplot(aes(x = species, fill = species)) +
  geom_bar() +
  labs(title = "Class Distribution", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")

# Correlation heatmap for numeric variables
numeric_data <- data %>% select_if(is.numeric)
cor_matrix <- cor(numeric_data)

p2 <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation") %>%
  ggplot(aes(var1, var2, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0) +
  labs(title = "Feature Correlations") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 + p2
```

## Train/Test Split

```{r data-split}
# Create stratified train/test split
set.seed(params$random_seed)
data_split <- initial_split(data, prop = 1 - params$test_prop, strata = species)
train_data <- training(data_split)
test_data <- testing(data_split)

cat("Training set:", nrow(train_data), "observations\n")
cat("Test set:", nrow(test_data), "observations\n")
cat("Class balance in training:")
table(train_data$species)
```

# Method 1: Principal Component Analysis (PCA)

## PCA: Classical Approach

**Principal Component Analysis (PCA)**:
- **Linear transformation** that maximizes variance
- Finds orthogonal directions (principal components)
- **Assumption**: Linear relationships capture the important structure
- **Output**: Coordinates in PC space

**Mathematical Foundation:**
$$\mathbf{Y} = \mathbf{X} \mathbf{W}$$
where $\mathbf{W}$ contains the eigenvectors of the covariance matrix.

## PCA Implementation

```{r pca-analysis}
# Prepare data for PCA (numeric variables only)
pca_recipe <- recipe(species ~ ., data = train_data) %>%
  step_select(all_numeric_predictors()) %>%
  step_normalize(all_numeric_predictors()) %>%
  step_pca(all_numeric_predictors(), num_comp = params$n_components)

# Fit PCA transformation
pca_prep <- prep(pca_recipe, training = train_data)

# Transform both training and test sets
train_pca <- bake(pca_prep, new_data = train_data)
test_pca <- bake(pca_prep, new_data = test_data)

# Extract PCA transformation details
pca_obj <- pca_prep$steps[[3]]  # PCA step
pca_loadings <- pca_obj$res$rotation[, 1:params$n_components]
pca_variance <- summary(pca_obj$res)$importance[2, 1:params$n_components]

cat("PCA Variance explained:")
print(pca_variance)
```

## PCA Visualization: Transformed Space

```{r pca-visualization, echo=FALSE, fig.height=5}
# Visualize PCA transformation
p1 <- train_pca %>%
  bind_cols(species = train_data$species) %>% 
  ggplot(aes(PC1, PC2, color = species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  labs(
    title = "PCA: Training Data in PC Space",
    x = paste0("PC1 (", round(pca_variance[1]*100, 1), "% variance)"),
    y = paste0("PC2 (", round(pca_variance[2]*100, 1), "% variance)")
  ) +
  theme_minimal()

p2 <- test_pca %>%
  bind_cols(species = test_data$species) %>% 
  ggplot(aes(PC1, PC2, color = species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  labs(
    title = "PCA: Test Data in PC Space",
    x = paste0("PC1 (", round(pca_variance[1]*100, 1), "% variance)"),
    y = paste0("PC2 (", round(pca_variance[2]*100, 1), "% variance)")
  ) +
  theme_minimal()

p1 + p2 + plot_layout(guides = "collect")
```

## PCA: Feature Loadings

```{r pca-loadings, echo=FALSE}
# Visualize PCA loadings
pca_loadings %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  pivot_longer(-feature, names_to = "component", values_to = "loading") %>%
  ggplot(aes(feature, loading, fill = component)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "PCA Feature Loadings",
    x = "Original Features",
    y = "Loading Coefficient"
  ) +
  theme_minimal()
```

# Method 2: Density Matrix (DM)

## DM: Quantum-Inspired Approach

**Density Matrix Method**:
- **Quantum-inspired transformation** via density matrices
- Creates non-linear coordinate system
- **Converts to polar coordinates**: $(r, \phi)$
- **No linearity assumption** - captures complex relationships

**Mathematical Foundation:**
$$\rho_d = \frac{XX^T}{\text{tr}(XX^T)}$$
Followed by SVD and coordinate transformation to polar space.

## DM Implementation

```{r dm-analysis}
# Fit DM model using the modern qml interface
dm_model <- dm_fit(species ~ ., data = train_data, n_breaks = params$n_breaks)

# Get model summary
print(dm_model)
```

## DM Implementation

```{r dm-analysis}
# Extract training data characteristics
cat("\nTraining data characteristics:")
for(i in seq_along(dm_model$trained_data)) {
  cat("\nClass", i, "(", dm_model$classes[i], "):", 
      nrow(dm_model$trained_data[[i]]), "observations")
}
```

## DM Visualization: Polar Coordinate Space

```{r dm-visualization, echo=FALSE, fig.height=5}
# Create visualization data by extracting polar coordinates from trained model
# Use the NEW feature: return_coords = TRUE to get test data coordinates

# Get training data polar coordinates
train_dm_viz <- bind_rows(
  dm_model$trained_data[[1]] %>% mutate(species = dm_model$classes[1]),
  dm_model$trained_data[[2]] %>% mutate(species = dm_model$classes[2])
)

# NEW FEATURE: Get test data with polar coordinates
test_pred_with_coords <- predict(dm_model, newdata = test_data, return_coords = TRUE)

# Extract test coordinates and add actual class labels
test_dm_viz <- test_pred_with_coords$coords %>%
  bind_cols(species = test_data$species)

# Training data visualization
p1 <- train_dm_viz %>%
  ggplot(aes(r, phi_1, color = species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  coord_polar() +
  labs(
    title = "DM: Training Data in Polar Space",
    x = "r (radial distance)",
    y = "φ₁ (angle)"
  ) +
  theme_minimal()

# Test data visualization - NOW POSSIBLE with return_coords = TRUE!
p2 <- test_dm_viz %>%
  ggplot(aes(r, phi_1, color = species)) +
  geom_point(size = 2.5, alpha = 0.7) +
  coord_polar() +
  labs(
    title = "DM: Test Data in Polar Space",
    x = "r (radial distance)", 
    y = "φ₁ (angle)"
  ) +
  theme_minimal()

p1 + p2 + plot_layout(guides = "collect")
```

## DM: Coordinate Space Properties

```{r dm-properties, echo=FALSE}
# Analyze the polar coordinate distributions for both training and test
train_coord_stats <- train_dm_viz %>%
  group_by(species) %>%
  summarise(
    dataset = "Training",
    n = n(),
    r_mean = mean(r, na.rm = TRUE),
    r_sd = sd(r, na.rm = TRUE),
    phi_mean = mean(phi_1, na.rm = TRUE),
    phi_sd = sd(phi_1, na.rm = TRUE),
    .groups = "drop"
  )

test_coord_stats <- test_dm_viz %>%
  group_by(species) %>%
  summarise(
    dataset = "Test",
    n = n(),
    r_mean = mean(r, na.rm = TRUE),
    r_sd = sd(r, na.rm = TRUE),
    phi_mean = mean(phi_1, na.rm = TRUE),
    phi_sd = sd(phi_1, na.rm = TRUE),
    .groups = "drop"
  )

coord_stats <- bind_rows(train_coord_stats, test_coord_stats) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

coord_stats %>%
  knitr::kable(caption = "Polar Coordinate Statistics by Species and Dataset")
```

## NEW FEATURE: Polar Coordinates Visualization

**The `predict()` function now supports `return_coords = TRUE`:**

```{r coords-feature-demo, echo=TRUE}
# Standard prediction
pred_standard <- predict(dm_model, test_data, type = "class")

# NEW: Predictions with polar coordinates
pred_with_coords <- predict(dm_model, test_data, return_coords = TRUE)

cat("Standard prediction structure:\n")
str(pred_standard)

cat("\nWith coordinates structure:\n") 
str(pred_with_coords)

cat("\nFirst few polar coordinates:\n")
head(pred_with_coords$coords)
```

# Classification Performance

## Metrics Setup

```{r metrics-setup}
# Define comprehensive metrics for comparison
classification_metrics <- metric_set(
  accuracy, bal_accuracy, sens, spec, 
  precision, recall, f_meas, kap, roc_auc
)
```

## PCA + Logistic Regression Performance

```{r pca-classification}
# Fit logistic regression on PCA-transformed data
pca_model <- logistic_reg() %>%
  set_engine("glm") %>%
  fit(species ~ ., data = train_pca %>% bind_cols(species = train_data$species))

# Make predictions
pca_pred <- predict(pca_model, test_pca, type = "class")
pca_prob <- predict(pca_model, test_pca, type = "prob")

# Calculate metrics
pca_results <- test_pca %>%
  bind_cols(species = test_data$species) %>%
  select(species) %>%
  bind_cols(pca_pred, pca_prob)

pca_metrics <- classification_metrics(pca_results, truth = species, estimate = .pred_class, .pred_Adelie)
```

## DM Classification Performance  

```{r dm-classification}
# Make predictions with DM model
dm_pred <- predict(dm_model, test_data, type = "class")
dm_prob <- predict(dm_model, test_data, type = "prob")

# Calculate metrics
dm_results <- test_data %>%
  select(species) %>%
  bind_cols(.pred_class = dm_pred) %>%
  bind_cols(.pred = dm_prob)

dm_metrics <- classification_metrics(dm_results, truth = species, estimate = .pred_class, .pred_Adelie)
```

## Performance Comparison Table

```{r performance-comparison, echo=FALSE}
# Combine metrics for comparison
comparison_metrics <- bind_rows(
  pca_metrics %>% mutate(method = "PCA + LogReg"),
  dm_metrics %>% mutate(method = "Density Matrix")
) %>%
  select(method, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

comparison_metrics %>%
  knitr::kable(caption = "Classification Performance Comparison")
```

## Confusion Matrices

```{r confusion-matrices, echo=FALSE, fig.height=4}
# PCA confusion matrix
pca_cm <- pca_results %>%
  conf_mat(truth = species, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "PCA + Logistic Regression") +
  theme_minimal()

# DM confusion matrix  
dm_cm <- dm_results %>%
  conf_mat(truth = species, estimate = .pred_class) %>%
  autoplot(type = "heatmap") +
  labs(title = "Density Matrix") +
  theme_minimal()

pca_cm + dm_cm
```

# Detailed Comparison

## Prediction Probability Distributions

```{r probability-distributions, echo=FALSE, fig.height=4}
# Compare prediction probability distributions
pca_prob_plot <- pca_results %>%
  select(species, .pred_Adelie, .pred_Gentoo) %>%
  pivot_longer(starts_with(".pred_"), names_to = "predicted_class", values_to = "probability") %>%
  mutate(predicted_class = str_remove(predicted_class, ".pred_")) %>%
  ggplot(aes(probability, fill = predicted_class)) +
  geom_histogram(alpha = 0.7, bins = 15) +
  facet_wrap(~species) +
  labs(title = "PCA: Prediction Probabilities", x = "Probability", y = "Count") +
  theme_minimal()

dm_prob_plot <- dm_results %>%
  select(species, .pred_Adelie, .pred_Gentoo) %>%
  pivot_longer(starts_with(".pred_"), names_to = "predicted_class", values_to = "probability") %>%
  mutate(predicted_class = str_remove(predicted_class, ".pred_")) %>%
  ggplot(aes(probability, fill = predicted_class)) +
  geom_histogram(alpha = 0.7, bins = 15) +
  facet_wrap(~species) +
  labs(title = "DM: Prediction Probabilities", x = "Probability", y = "Count") +
  theme_minimal()

pca_prob_plot / dm_prob_plot
```

## ROC Curves Comparison

```{r roc-comparison, echo=FALSE, fig.height=4}
# ROC curves
pca_roc <- pca_results %>%
  roc_curve(truth = species, .pred_Adelie) %>%
  mutate(method = "PCA")

dm_roc <- dm_results %>%
  roc_curve(truth = species, .pred_Adelie) %>%
  mutate(method = "DM")

bind_rows(pca_roc, dm_roc) %>%
  ggplot(aes(1 - specificity, sensitivity, color = method)) +
  geom_line(size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(
    title = "ROC Curve Comparison",
    x = "1 - Specificity",
    y = "Sensitivity"
  ) +
  theme_minimal() +
  coord_equal()
```

## Computational Characteristics

```{r computational-comparison, echo=FALSE}
# Compare computational aspects
comp_comparison <- tribble(
  ~Aspect, ~PCA, ~DM,
  "Transformation Type", "Linear", "Non-linear (Quantum-inspired)",
  "Output Space", "Principal Components", "Polar Coordinates (r, φ)",
  "Interpretability", "High (variance explained)", "Moderate (quantum matrix)",
  "Scalability", "Excellent", "Good",
  "Parameter Tuning", "# Components", "# Breaks, bandwidth",
  "Preprocessing", "Normalization required", "Handles mixed data types",
  "Assumptions", "Linear relationships", "Minimal assumptions"
)

comp_comparison %>%
  knitr::kable(caption = "Methodological Comparison")
```

## Summary: When to Use Each Method

**PCA is preferred when:**
- **Linear relationships** dominate the data structure
- **Interpretability** is crucial (variance explained)
- **Large datasets** require fast computation
- **Feature reduction** is the primary goal

**Density Matrix is preferred when:**
- **Non-linear patterns** are suspected
- **Complex data relationships** need to be captured
- **Mixed data types** (continuous + categorical)
- **Novel quantum-inspired approaches** are of interest

# Conclusions

## Key Findings

```{r key-findings, echo=FALSE}
# Extract key performance differences
accuracy_diff <- comparison_metrics$accuracy[comparison_metrics$method == "Density Matrix"] - 
                 comparison_metrics$accuracy[comparison_metrics$method == "PCA + LogReg"]

f1_diff <- comparison_metrics$f_meas[comparison_metrics$method == "Density Matrix"] - 
           comparison_metrics$f_meas[comparison_metrics$method == "PCA + LogReg"]

auc_diff <- comparison_metrics$roc_auc[comparison_metrics$method == "Density Matrix"] - 
            comparison_metrics$roc_auc[comparison_metrics$method == "PCA + LogReg"]
```

**Performance Summary:**
- **Accuracy difference**: `r round(accuracy_diff, 3)` (DM vs PCA)
- **F1-score difference**: `r round(f1_diff, 3)` (DM vs PCA)  
- **AUC difference**: `r round(auc_diff, 3)` (DM vs PCA)

**Method Characteristics:**
- **PCA**: Fast, interpretable, linear assumptions
- **DM**: Novel approach, handles non-linearity, quantum-inspired

## Future Directions

**Research Opportunities:**
1. **Scalability studies** with larger datasets
2. **Multi-class extension** for DM method
3. **Hybrid approaches** combining PCA and quantum methods
4. **Theoretical analysis** of quantum-classical performance gaps

**Implementation Improvements:**
1. **Full test data visualization** in polar coordinates
2. **Cross-validation** comparison framework  
3. **Feature importance** analysis for DM
4. **Computational efficiency** optimization

## References and Acknowledgments

**Quantum Machine Learning Package:**
- Modern `dm_fit()` and `predict()` interface
- Follows tidymodels conventions
- Available at: https://github.com/RodGal-2020/qml

**Dataset:**
- Palmer Penguins dataset via `modeldata` package
- Binary classification: Adelie vs Gentoo penguins

**Tools:**
- R tidymodels ecosystem
- ggplot2 for visualizations  
- Custom quantum matrix implementation
