---
title: "QML Package Testing and Validation"
author: "Xopre Rodríguez Gallego"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    highlight: tango
params:
  fast_test: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  error = TRUE,
  collapse = TRUE,
  comment = "#>"
)
```

# QML Package This document provides comprehensive testing and validation of the `qml` package (Quantum Machine Learning). It covers package installation, basic functionality, advanced features, and performance validation.
d performance validation.

## 1. Package Development Setup

### 1.1 Update Package Documentation

```{r update-docs}
# Update documentation and NAMESPACE
if (requireNamespace("devtools", quietly = TRUE)) {
  cat("Updating package documentation...\n")
  devtools::document()
  cat("✓ Documentation updated\n")
} else {
  cat("devtools not available - install with: install.packages('devtools')\n")
}
```

### 1.2 Install Package from Source

```{r install-package}
if (params$fast_test) {
  cat("⚠ Fast testing mode enabled - skipping package installation\n")
  source("../R/dm_utils.R")
  source("../R/data_preprocessing.R")  
  source("../R/quantum_matrix.R")
  source("../R/kernel_functions.R")
  source("../R/coordinate_transforms.R")
  source("../R/dm_classifier.R")
} else {
  # Install the package from the current directory
  if (requireNamespace("devtools", quietly = TRUE)) {
    cat("Installing qml package from source...\n")
    try({
      devtools::install(".", upgrade = "never", quiet = TRUE)
      cat("✓ Package installed successfully\n")
    })
  } else {
    cat("⚠ devtools not available - loading source files directly\n")
    # Fallback: source files directly
    source("../R/dm_utils.R")
    source("../R/data_preprocessing.R")  
    source("../R/quantum_matrix.R")
    source("../R/kernel_functions.R")
    source("../R/coordinate_transforms.R")
    source("../R/dm_classifier.R")
  }
}
```

### 1.3 Load Required Libraries

```{r load-libraries}
# Load required packages
required_packages <- c("dplyr", "magrittr", "purrr", "tibble")

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    cat("Installing", pkg, "...\n")
    install.packages(pkg)
  }
  library(pkg, character.only = TRUE)
}

# Try to load the qml package
if (!params$fast_test) {
  if (requireNamespace("qml", quietly = TRUE)) {
    library(qml)
    cat("✓ qml package loaded successfully\n")
  } else {
    cat("⚠ qml package not available as installed package\n")
  }
}
```

## 2. Package Structure Validation

### 2.1 Check Package Information

```{r package-info}
# Check if package is properly loaded
if (exists("dm_fit")) {
  cat("✓ Main functions available\n")
  
  # List exported functions
  if (requireNamespace("qml", quietly = TRUE)) {
    exports <- ls("package:qml")
    # cat("Exported functions:", paste(exports, collapse = ", "), "\n")
  }
} else {
  cat("⚠ Main functions not found\n")
}

# Check key functions exist
key_functions <- c("dm_fit", "K", "f_hat_h")
for (func in key_functions) {
  if (exists(func)) {
    cat("✓", func, "available\n")
  } else {
    cat("✗", func, "missing\n")
  }
}
```

### 2.2 Validate Dependencies

```{r check-dependencies}
# Check if all required packages are available
deps <- c("dplyr", "magrittr", "purrr", "tibble", "rlang", "fasano.franceschini.test")

cat("Checking dependencies:\n")
for (dep in deps) {
  if (requireNamespace(dep, quietly = TRUE)) {
    cat("✓", dep, "available\n")
  } else {
    cat("✗", dep, "missing - install with: install.packages('", dep, "')\n")
  }
}
```

## 3. Basic Functionality Testing

### 3.1 Data Preparation

```{r data-prep}
# Load test dataset
data(iris)
cat("Dataset loaded: iris\n")
cat("Dimensions:", nrow(iris), "rows,", ncol(iris), "columns\n")
cat("Classes:", paste(levels(iris$Species), collapse = ", "), "\n")

# Check data structure
str(iris)
head(iris)
```

### 3.3 Test Kernel Function

```{r test-kernel}
cat("Testing K() kernel function:\n")

if (exists("K")) {
  try({
    # Test various kernel values
    test_cases <- list(
      c(0, 0),      # Origin
      c(0.5, 0.5),  # Inside unit circle
      c(1, 0),      # On boundary
      c(1.5, 0)     # Outside boundary
    )
    
    for (i in seq_along(test_cases)) {
      coords <- test_cases[[i]]
      k_val <- K(coords[1], coords[2], r = 2)
      cat("K(", coords[1], ",", coords[2], ") =", k_val, "\n")
    }
    
    cat("✓ K() function working correctly\n")
    
  }, silent = FALSE)
} else {
  cat("✗ K() function not available\n")
}
```

## 4. Model Fitting and Prediction Testing

### 4.1 Basic Model Fitting

```{r test-model-fitting}
cat("Testing dm_fit() function:\n")

filtered_iris <- iris %>% dplyr::filter(Species != "setosa")  # Binary classification for simplicity

if (exists("dm_fit")) {
  try({
    # Fit basic model
    model <- dm_fit(Species ~ ., data = filtered_iris, n_breaks = 3, verbose = 1)
    
    cat("✓ Model fitted successfully\n")
    cat("Model class:", class(model), "\n")
    cat("Number of classes:", model$n_classes, "\n")
    cat("Classes:", paste(model$classes, collapse = ", "), "\n")
    
    # Print model summary
    print(model)
    
  }, silent = FALSE)
} else {
  cat("✗ dm_fit() function not available\n")
}
```

### 4.2 Model Prediction Testing

```{r test-predictions}
if (exists("model") && inherits(model, "dm_fit")) {
  cat("Testing predict() methods:\n")
  
  try({
    # Test class predictions
    pred_class <- predict(model, newdata = iris[1:10, ])
    cat("✓ Class predictions successful\n")
    cat("Predictions:", as.character(pred_class), "\n")
    cat("Actual:", as.character(iris$Species[1:10]), "\n")
    
    # Test probability predictions
    pred_prob <- predict(model, newdata = iris[1:5, ], type = "prob")
    cat("✓ Probability predictions successful\n")
    cat("Probability matrix dimensions:", dim(pred_prob), "\n")
    print(round(pred_prob, 3))
    
    # Check probabilities sum to 1
    row_sums <- rowSums(pred_prob)
    cat("Row sums (should be ~1):", round(row_sums, 3), "\n")
    
  }, silent = FALSE)
} else {
  cat("⚠ Model not available for prediction testing\n")
}
```

### 4.3 Model Summary Testing

```{r test-summary}
if (exists("model") && inherits(model, "dm_fit")) {
  cat("Testing summary() method:\n")
  
  try({
    summary(model)
    cat("✓ Summary method working correctly\n")
  }, silent = FALSE)
} else {
  cat("⚠ Model not available for summary testing\n")
}
```

## 5. Advanced Testing

### 5.1 Train/Test Split Validation

```{r advanced-testing}
cat("Advanced testing with train/test split:\n")

if (exists("testify") && exists("dm_fit")) {
  try({
    # Create proper train/test split
    iris_split <- testify(iris, "Species", test_prob = 0.3)
    train_data <- iris_split[!iris_split$test, ]
    test_data <- iris_split[iris_split$test, ]
    
    cat("Training set size:", nrow(train_data), "\n")
    cat("Test set size:", nrow(test_data), "\n")
    
    # Fit model on training data
    model_split <- dm_fit(Species ~ ., data = train_data, n_breaks = 3)
    
    # Predict on test data
    test_predictions <- predict(model_split, newdata = test_data)
    test_probabilities <- predict(model_split, newdata = test_data, type = "prob")
    
    # Create confusion matrix
    conf_matrix <- table(Predicted = test_predictions, Actual = test_data$Species)
    cat("\nConfusion Matrix:\n")
    print(conf_matrix)
    
    # Calculate accuracy
    accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)
    cat("Accuracy:", round(accuracy * 100, 2), "%\n")
    
    # Calculate per-class accuracy
    class_accuracy <- diag(conf_matrix) / rowSums(conf_matrix)
    names(class_accuracy) <- rownames(conf_matrix)
    cat("Per-class accuracy:\n")
    print(round(class_accuracy * 100, 2))
    
    cat("✓ Advanced testing completed successfully\n")
    
  }, silent = FALSE)
} else {
  cat("⚠ Required functions not available for advanced testing\n")
}
```

### 5.2 Parameter Sensitivity Testing

```{r parameter-testing}
cat("Testing different parameter values:\n")

if (exists("dm_fit")) {
  try({
    # Test different n_breaks values
    n_breaks_values <- c(2, 3, 5)
    results <- list()
    
    for (nb in n_breaks_values) {
      cat("Testing n_breaks =", nb, ":\n")
      
      model_test <- dm_fit(Species ~ ., data = iris[1:100, ], n_breaks = nb)
      pred_test <- predict(model_test, newdata = iris[101:150, ])
      
      # Simple accuracy on remaining data
      actual <- iris$Species[101:150]
      accuracy <- mean(pred_test == actual)
      
      results[[paste0("n_breaks_", nb)]] <- accuracy
      cat("  Accuracy:", round(accuracy * 100, 2), "%\n")
    }
    
    cat("Parameter testing results:\n")
    print(results)
    
  }, silent = FALSE)
} else {
  cat("⚠ dm_fit() not available for parameter testing\n")
}
```

## 6. Error Handling Testing

### 6.1 Input Validation Testing

```{r error-testing}
cat("Testing error handling:\n")

if (exists("dm_fit")) {
  # Test missing formula
  cat("1. Testing missing formula:\n")
  try({
    dm_fit()
  }, silent = TRUE)
  
  # Test invalid data
  cat("2. Testing with insufficient data:\n")
  try({
    small_data <- iris[1:6, ]  # Only 2 observations per class
    dm_fit(Species ~ ., data = small_data)
  }, silent = TRUE)
  
  # Test missing objective variable
  cat("3. Testing missing objective variable:\n")
  try({
    iris_no_species <- iris[, -5]
    dm_fit(Species ~ ., data = iris_no_species)
  }, silent = TRUE)
  
  cat("✓ Error handling tests completed\n")
} else {
  cat("⚠ dm_fit() not available for error testing\n")
}
```

## 7. Performance Benchmarking

### 7.1 Timing Analysis

```{r performance-testing}
cat("Performance benchmarking:\n")

if (exists("dm_fit")) {
  try({
    # Time model fitting
    fit_time <- system.time({
      perf_model <- dm_fit(Species ~ ., data = iris, n_breaks = 3)
    })
    
    cat("Model fitting time:", round(fit_time["elapsed"], 3), "seconds\n")
    
    # Time predictions
    pred_time <- system.time({
      perf_pred <- predict(perf_model, newdata = iris)
    })
    
    cat("Prediction time (150 obs):", round(pred_time["elapsed"], 3), "seconds\n")
    cat("Prediction rate:", round(nrow(iris) / pred_time["elapsed"], 0), "observations/second\n")
    
  }, silent = FALSE)
} else {
  cat("⚠ Functions not available for performance testing\n")
}
```

## 8. Package Validation Summary

### 8.1 Test Results Summary

```{r test-summary}
cat("QML Package Validation Summary\n")
cat("==============================\n\n")

tests_passed <- 0
tests_total <- 8

# Check each component
components <- list(
  "Package Loading" = exists("dm_fit"),
  "Data Splitting (testify)" = exists("testify"),
  "Kernel Function (K)" = exists("K"),
  "Model Fitting (dm_fit)" = exists("dm_fit"),
  "Predictions" = exists("model") && inherits(model, "dm_fit"),
  "Summary Methods" = exists("model") && inherits(model, "dm_fit"),
  "Advanced Features" = exists("testify") && exists("dm_fit"),
  "Error Handling" = exists("dm_fit")
)

for (name in names(components)) {
  if (components[[name]]) {
    cat("✓", name, "\n")
    tests_passed <- tests_passed + 1
  } else {
    cat("✗", name, "\n")
  }
}

cat("\nOverall Test Results:", tests_passed, "/", tests_total, "passed\n")

if (tests_passed == tests_total) {
  cat("🎉 ALL TESTS PASSED! Package is ready for use.\n")
} else if (tests_passed >= tests_total * 0.75) {
  cat("⚠ Most tests passed. Package is mostly functional.\n")
} else {
  cat("❌ Many tests failed. Package needs attention.\n")
}
```

### 8.2 Recommendations

```{r recommendations}
cat("\nRecommendations:\n")
cat("================\n\n")

if (tests_passed == tests_total) {
  cat("✓ Package is functioning correctly\n")
  cat("✓ Ready for distribution\n")
  cat("✓ Consider adding more comprehensive tests\n")
  cat("✓ Add performance benchmarks for larger datasets\n")
} else {
  cat("• Install missing dependencies\n")
  cat("• Fix any errors shown above\n")
  cat("• Run devtools::check() for comprehensive validation\n")
  cat("• Consider running tests in clean R session\n")
}

cat("\nNext Steps:\n")
cat("• Run devtools::check() for CRAN readiness\n")
cat("• Build package documentation with devtools::build_manual()\n")
cat("• Test with different R versions if needed\n")
cat("• Submit to CRAN when ready\n")
```

---

**Package Testing Complete**

*Generated on `r Sys.time()`*
